Compositional Quantitative Resource Analysis of C Programs

### 2014-06-13 paper/main.tex
done with the catalog for now

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use cost metric that counts cost $1$ for assignments
and cost $0$ for all other operations.  Of course, the examples can
also be analyzed with any other cost metric.

We assume that the variables in the code snippets are the inputs to
the program. Some of the examples contain constants on which the
computed bound depends.  These constants are randomly chosen to
present an example but the analysis works for other constants as well.
Note however that it is sometimes crucial that constants are positive
(or negative) and that the same constant at different places.

@@

   \caption{Examples from Gulwani et al's SPEED (a)~\cite{GulwaniMC09}}
  \label{fig:cat2a}
\end{figure*}

@@

   \caption{Examples from Gulwani et al's SPEED (b)~\cite{GulwaniMC09}}
  \label{fig:cat2b}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$4 + 4|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$1 + 1|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig\_4\_5}
    \end{center}
  \end{minipage}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i; j=j+1;
  }
  while (j<n) {
    if (0) {
      _tick(1);
      j=j-1; n=n-1;
    }
    j=j+1;
  }
  i=i+1;
   \end{lstlisting}
$1 + |[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    }
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (0) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$2 |[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$2 + 3|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}

@@

Our analyzer can derive almost all bounds for programs that have been
described as challenges in the literature on bound generation.  There
are only two types of programs that we were not able to handle:
programs that require \emph{super-linear bounds} and program that
require \emph{path-sensitive reasoning} to derive a bound.  In the
following we list all linear examples from the literature that we
found.  Two of the require path-sensitive reasoning.

@@

>>>>

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use cost metric that counts cost $1$ for assignments
and cost $0$ for all other operations.  Sometimes we also use the
\emph{ticks metric} if we want to discuss features like returning
resources.  Of course, the examples can also be analyzed with any other
cost metric.

We assume that free variables in the code snippets are the inputs to
the program. Some of the examples contain constants on which the
computed bound depends.  These constants are randomly chosen to
present an example but the analysis works for other constants as well.
Note however that it is sometimes crucial that constants are positive
(or negative) and that the same constant at different places.

@@

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09}} (a)
  \label{fig:cat2a}
\end{figure*}

@@

   \caption{Examples from Gulwani et al's SPEED~\cite{GulwaniMC09} (b)}
  \label{fig:cat2b}
\end{figure*}

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$2 + 4|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig\_4\_5}
    \end{center}
  \end{minipage}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  i=0;
  while (i<n) {
    j=i+1;
    while (j<n) {
      if (*) {
	_tick(1);
	j=j-1; n=n-1;
      }
      j=j+1;
    }
    i=i+1;
  }
   \end{lstlisting}
$|[0, n]| \text{ ticks}$
\\[.7\baselineskip]
      {\bf ex1}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0 && m>0) {
    n--; m--;
    while (nondet()) {
      n--; m++;
    }
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf ex2}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (n>0) {
    t = x;
    n=n-1;
    while (n>0) {
      if (*) break;
      n=n-1;
    }
  }
   \end{lstlisting}
$2 |[0, n]|$
\\[.7\baselineskip]
      {\bf ex3}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  flag=1;
  while (flag>0) {
    flag=0;
    while (n>0) {
      n=n-1;       
      flag=1;
    }
  }
   \end{lstlisting}
$2 + 3|[0, n]|$
\\[.7\baselineskip]
      {\bf ex4}
    \end{center}
  \end{minipage}

@@

Our analyzer can derive almost all linear bounds for programs that
have been described as challenges in the literature on bound
generation.  We found only three programs with a linear bound for
which our analyzer could not find a tight bound.  The first one
(\emph{fig\_4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
  reasoning} to derive a bound.  The other two, \emph{ex2} and
\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
to have a linear bound but seem to be non-terminating.

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani' et
al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove the termination.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

Example \emph{nested\_multiple} is similar to example \emph{fig2\_1}.
Instead of incrementing variable $y$ in the outer loop, $y$ is here
potentially incremented multiple times in each iteration of the outer
loop.  The idea of example \emph{nested\_single} is similar.  However,
instead of incrementing variable $y$ in the inner loop, we increment
$x$, the counter variable of the outer loop. Our analyzer derives a
tight bound for both programs.  Note that a star \emph{*} in a
branching condition denotes an arbitrary boolean condition that might
of course change while iterating (non-deterministic choice).

Example \emph{sequential\_single} is like example
\emph{nested\_single}.  The only difference is that the inner loop of
\emph{nested\_single} is now evaluated after the outer loop.  Example
\emph{simple\_multiple} is a variant of example \emph{fig2\_1} and
\emph{simple\_single} is a simple variant of \emph{nested\_single}.
We derive tight bounds for all aforementioned programs.

Example \emph{simple\_single\_2} uses conditionals and a \code{break}
statement to control loop iterations.  If $x<N$ then variables $x$ and
$y$ are incremented.  Otherwise, if $y<M$ then the same increment is
executed.  If $y\geq M$ and $x\geq N$ then the loop is terminated with
a break.  Our tool computes the bound $2 + 2 |[0, M]| + 2 |[0, N]|$.
This bound is tight in the sense that there are inputs (such as $M =
-100$ and $N = 100$) for which the bound precisely describes the
execution cost.  However, SPEED can compute the more precise bound
$\max(N,M)$.  We currently cannot express this bound in our system.

Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$) and the potential of $|0,vb|$ ($2$).  Since we also
know that $|[0, n]|>0$ we obtain constant potential $4$ and establish
the loop invariant again.  In the \emph{else branch}, we use the
potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$ potential
units to pay for the assignment.  Since we have $4$ constant potential
units left after the loop, we can subtract them from the initial
potential $6 + 4|[0, n]|$.

In example \emph{fig\_4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the else branch of the conditional since $|[0,i]|$
decreases.  Example \emph{fig\_4\_5} is one of the three examples that
we can not handle automatically.  The reasoning is based on the fact
that the boolean value of the test \code{dir==fwd} does not change
during the iteration of the loop.  As a result, the variable $i$ is
either counted down to $0$ or up to $n$.  Our tool cannot handle
example \emph{fig\_4\_5} because we don't do path sensitive reasoning.
Note however that it would be more efficient to move the test
\code{dir==fwd} outside of the loop (this would be also done by an
optimizing compiler).  The resulting program can be analyzer by our
tool.

Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
code in \emph{if} statement.  So we use the \emph{tick metric} and
insert \code{_tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
$n=n-1$.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.

We cannot analyze example \emph{ex2} for which previous work reported
a linear bound [GulwaniPLDI10].  The reason is that this example is
not terminating since the inner loop is not bounded.  The same is true
for example \emph{ex5} from the same paper, which is not listed here.
Example \emph{ex2} is similar to example \emph{nested\_single} and
\emph{ex3} is a variant of example \emph{t47}.

@@

Our approach can naturally deal with mutually-recursive functions.
The recursion patterns can be exactly the same that are used in
iterations of loops.  In the following, we present three simple
examples that illustrate the analysis of functions.

Example \emph{t37} illustrates that the analyzer is able to perform
inter-procedural size tracking.  The function \code{copy} adds the
argument $x$ to the argument $y$ if $x$ is positive.  However, this
addition is done in steps of $1$ in each recursive call.  The function
\code{count\_down} recursively decrements its argument down to $0$.
The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
\code{_main} in which we first add $x$ to $y$ using the function
\code{copy} and then count down the variable $y$ using the function
\code{count\_down}.  The derived bound is tight.

Example \emph{t39} uses mutual recursion.  The function
\code{count\_down} is similar to the function with the same name in
example \emph{t37}.  However, we do not count down to $0$ but to a
variable $y$ that is passed as an argument and we call the function
\code{count\_up} afterwards.  The function \code{count\_up} is dual to
\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
\code{count\_down}.  For the function $\_main$ that calls
\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
1.33 |[z,y]|$.

Example \emph{t46} shows a program that uses and returns resources.
Again, we use the \emph{tick metric} and the function \code{\_tick} to
describe the resource usage.  The function \code{produce} produces
$|[0,x]|$ resources, that is, in each of the $|[0,x]|$ iterations, it
receives one resource unit.  Similarly, the function \code{consume}
consumes $|[0,y]|$ resources.  The analyzer computes the tight bound
$|[0,y]|$ for the function \code{\_main}.  This is only possible since
amortized analysis naturally tracks the size changes to the variables
$x$ and $y$, and the interaction between \code{consume} and
\code{produce}.
